Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimization Choices

### 2025-11-20 jso.qmd
change color palette

```{r}
#| label: fig-squintability-ill
#| fig.width: 8
#| fig.height: 4
#| fig.cap: "Simulated traces of index value as a function of projection distances for four PP index functions: MIC, splines2d, skinny, and stringy2. The index function MIC and splines2d make early progression towards the optimal index value, indicating a large squint angle, whereas the traces from skinny, and specially stringy2, show improvement only near the optimal projection, suggesting low squintability."
sq_sine_basis_df |> 
  filter(n == 4) |> 
  unnest(basis_df, names_repair = "unique") |> 
  filter(id <= 3) |> 
  rename(index = index...2, value = index...8) |> 
  filter(index %in% c("MIC", "skinny", "stringy2", "splines2d")) |> 
  group_by(id, index, n) |> 
  mutate(value = (value - min(value)) / (max(value) - min(value))) |>
  ggplot(aes(x = dist, y = value, 
             group = interaction(index, id, n), 
             color = index)) +
  geom_line() + 
  scale_color_brewer(palette = "Dark2") +
  xlab("Projection distance") + ylab("Index value") +
  theme_bw() + 
  theme(legend.position = "bottom")
```

@@

res_quantile |> 
  ungroup() |> 
  ggplot(aes(x = n_jellies - 0.5, y = proportion, 
             group = d, color = d)) + 
  geom_segment(aes(xend = n_jellies + 0.5), size = 0.05) + 
  geom_segment(aes(x = n_jellies - 1, xend = n_jellies + 1), 
               data = res_quantile |> summarise(proportion = mean(proportion)), 
               size = 2) + 
  geom_line(aes(x = n_jellies), 
               data = res_quantile |> summarise(proportion = mean(proportion)), 
            alpha = 0.3) + 
  scale_color_brewer(palette = "Dark2", name = "dimension") +
  scale_x_continuous(breaks = c(20, 50, 100), labels = c("20", "50", "100")) +
  facet_wrap(vars(`max # of iteration`), 
             labeller = "label_both") + 
  theme_bw() +
  theme(panel.grid.minor = element_blank()) + 
  xlab("Number of jellyfish") + 
  ylab("Proportion of success")
```

@@

```{r}
#| label: fig-idx-proj-dist
#| fig.width: 8
#| fig.height: 4
#| fig.cap: "Five traces of index value against projection distance for each of the three indexes (TIC, spline, and skinny) on 6D sine wave data. The traces of the splines index are smooth, with a gradual change between noise and otimal projection. In contrast, the skinny index has a noisy trace that increases rapidly near the optimum. This illustrates the usefulness of the new measures for squintability and smoothness."
sq_basis_df_6d_TIC_spline_skinny |>
  ggplot(aes(x = dist, y = index, group = interaction(index_name, id), color = index_name)) +
  geom_line() + 
  scale_color_brewer(palette = "Dark2", name = "Index function") +
  xlab("Projection distance") + ylab("Index value") +
  theme_bw() + 
  theme(legend.position = "bottom")
```

@@

The R packages used in this work include: `tidyr` [@tidyr], `dplyr`
[@dplyr], `ggplot2` [@ggplot2], `knitr` [@knitr],`gtsummary`
[@gtsummary], `patchwork` [@patchwork], `ggh4x` [@ggh4x], `broom`
[@broom], `kableExtra` [@kabelextra], `ferrn` [@RJ-2021-105], and
`cassowaryr` [@cassowaryr].

>>>>

```{r}
#| label: fig-squintability-ill
#| fig.width: 8
#| fig.height: 4
#| fig.cap: "Simulated traces of index value as a function of projection distances for four PP index functions: MIC, splines2d, skinny, and stringy2. The index function MIC and splines2d make early progression towards the optimal index value, indicating a large squint angle, whereas the traces from skinny, and specially stringy2, show improvement only near the optimal projection, suggesting low squintability."
sq_sine_basis_df |> 
  filter(n == 4) |> 
  unnest(basis_df, names_repair = "unique") |> 
  filter(id <= 3) |> 
  rename(index = index...2, value = index...8) |> 
  filter(index %in% c("MIC", "skinny", "stringy2", "splines2d")) |> 
  group_by(id, index, n) |> 
  mutate(value = (value - min(value)) / (max(value) - min(value))) |>
  ggplot(aes(x = dist, y = value, 
             group = interaction(index, id, n), 
             color = index)) +
  geom_line() + 
  scale_color_manual(values = palette.colors()[1:4]) + 
  xlab("Projection distance") + ylab("Index value") +
  theme_bw() + 
  theme(legend.position = "bottom")
```

@@

res_quantile |> 
  ungroup() |> 
  ggplot(aes(x = n_jellies - 0.5, y = proportion, 
             group = d, color = d)) + 
  geom_segment(aes(xend = n_jellies + 0.5), size = 0.05) + 
  geom_segment(aes(x = n_jellies - 1, xend = n_jellies + 1), 
               data = res_quantile |> 
                 summarise(proportion = mean(proportion)), 
               size = 2) + 
  geom_line(aes(x = n_jellies), 
            data = res_quantile |> 
              summarise(proportion = mean(proportion)), 
            alpha = 0.3) + 
  scale_color_manual(values = palette.colors()[1:5], name = "Dimension") + 
  scale_x_continuous(breaks = c(20, 50, 100), labels = c("20", "50", "100")) +
  facet_wrap(vars(`max # of iteration`), 
             labeller = "label_both") + 
  theme_bw() +
  theme(panel.grid.minor = element_blank()) + 
  xlab("Number of jellyfish") + 
  ylab("Proportion of success")
```

@@

```{r}
#| label: fig-idx-proj-dist
#| fig.width: 8
#| fig.height: 4
#| fig.cap: "Five traces of index value against projection distance for each of the three indexes (TIC, spline, and skinny) on 6D sine wave data. The traces of the splines index are smooth, with a gradual change between noise and otimal projection. In contrast, the skinny index has a noisy trace that increases rapidly near the optimum. This illustrates the usefulness of the new measures for squintability and smoothness."
sq_basis_df_6d_TIC_spline_skinny |>
  ggplot(aes(x = dist, y = index, group = interaction(index_name, id), color = index_name)) +
  geom_line() + 
  scale_color_manual(values = palette.colors()[1:3],
                     name = "Index function") + 
  #scale_color_brewer(palette = "Dark2", name = "Index function") +
  xlab("Projection distance") + ylab("Index value") +
  theme_bw() + 
  theme(legend.position = "bottom")
```

@@

The R packages used in this work include: `tidyr` [@tidyr], `dplyr`
[@dplyr], `ggplot2` [@ggplot2], `knitr` [@knitr], `gtsummary`
[@gtsummary], `patchwork` [@patchwork], `ggh4x` [@ggh4x], `broom`
[@broom], `kableExtra` [@kabelextra], `ferrn` [@RJ-2021-105], and
`cassowaryr` [@cassowaryr].

### 2025-11-20 jso.tex
change color palette

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, bottomtitle=1mm, colback=white, toprule=.15mm, left=2mm, opacityback=0, colframe=quarto-callout-note-color-frame, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, breakable, coltitle=black, toptitle=1mm, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, opacitybacktitle=0.6]

@@

The R packages used in this work include: \texttt{tidyr} (Wickham,
Vaughan, and Girlich 2024), \texttt{dplyr} (Wickham et al. 2023),
\texttt{ggplot2} (Wickham 2016), \texttt{knitr} (Xie
2014),\texttt{gtsummary} (Sjoberg et al. 2021), \texttt{patchwork}
(Pedersen 2024), \texttt{ggh4x} (Van Den Brand 2024), \texttt{broom}
(Robinson, Hayes, and Couch 2024), \texttt{kableExtra} (Zhu 2024),
\texttt{ferrn} (Zhang et al. 2021), and \texttt{cassowaryr} (Mason et
al. 2022).

>>>>

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, opacitybacktitle=0.6, left=2mm, toprule=.15mm, colback=white, leftrule=.75mm, arc=.35mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomtitle=1mm, rightrule=.15mm, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0]

@@

The R packages used in this work include: \texttt{tidyr} (Wickham,
Vaughan, and Girlich 2024), \texttt{dplyr} (Wickham et al. 2023),
\texttt{ggplot2} (Wickham 2016), \texttt{knitr} (Xie 2014),
\texttt{gtsummary} (Sjoberg et al. 2021), \texttt{patchwork} (Pedersen
2024), \texttt{ggh4x} (Van Den Brand 2024), \texttt{broom} (Robinson,
Hayes, and Couch 2024), \texttt{kableExtra} (Zhu 2024), \texttt{ferrn}
(Zhang et al. 2021), and \texttt{cassowaryr} (Mason et al. 2022).
